---
title: "Tidy IMF WEO - October 2022"
author: "Teal Emery"
format: gfm
editor: visual
---

# Introduction

This GitHub Repo houses a [tidy (long-format)](https://r4ds.hadley.nz/data-tidy.html) version of the [IMF's World Economic Outlook Database for October 2022](https://www.imf.org/en/Publications/SPROLLs/world-economic-outlook-databases#sort=%40imfdate%20descending). I use this a lot for data analysis and data visualization, and I want to make sure my work is reproducible, so I'm putting it on GitHub. Also, I hope the data or the scripts can be useful for others, especially those working on projects with me.

Someday I'll clean up & formalize the scripts, but they work well for now.

Elements of the Tidy WEO:

-   **Standardized Country Names**: The country names are standardized using the [countrycode R package](https://github.com/vincentarelbundock/countrycode). The dataset has standardized country names in the `country_name` column, and [ISO3C codes](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3) in the `iso3c` column.

-   **Analysis-Friendly Subject Names**: I added sensible names that are easy to filter and can easily be used in tables + data visualizations. You can see the subject names in the `codebook.csv` file

-   **Tidy Long-format**: Simplifies data analysis and data visualization, especially if you're a [R Tidyverse](https://www.tidyverse.org/) zealot, like I am.

## Data

The data is available in long-format as an `.rds` (R data file).

Long-format `.csv` files aren't space efficient, so I have it written as a wide-format `.csv` file. Making it wide-format cuts the file size down from 42.1 MB to 1.5 MB. I make this available for non-R users.

# Downloading the Data

if you're using R, you can copy the code for the function below to download the data.

```{r}
download_tidy_weo_2022_oct <- function() {
  
  # rds file github url
  github_url <- "https://github.com/t-emery/tidy_imf_weo_2022_oct/raw/master/00_data_processed/imf_weo_2022_oct_by_country_tidy.rds"
  
  github_url |> 
    url() |> 
    #unzips it
    gzcon() |> 
    #reads it
    readRDS()
}
```

You can run it like this:

```{r}
download_tidy_weo_2022_oct() |> 
  head()
```

To download the codebook:

```{r}
download_tidy_weo_2022_oct_codebook <- function() {
  
  # rds file github url
  github_url <- "https://raw.githubusercontent.com/t-emery/tidy_imf_weo_2022_oct/master/00_data_processed/codebook.csv"
  
  github_url |> 
    readr::read_csv()
}
```

```{r}
#| message: false
download_tidy_weo_2022_oct_codebook()
```
